Double Standards in Judging Collective Action\thanks{}

### 2023-04-26 Documents/manuscript.qmd
add paragraph on partisanship as a social identity

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In five studies (total *N* = 2,979), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In three preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own ideological position---but showed no consistent ingroup bias when judging collective action by ingroup and outgroup members. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
keywords: collective action, double standards, social identity theory, system justification theory, Black Lives Matter
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: apa-template.tex
---

@@

<!-- Write it in your own words and structure it (with references) later. -->
<!-- Political orientation can be understood as ideology and identity. -->

- Are Partisan Double Standards Based in Ideology or Identity?
- Evidence for partisanship as a elective affinity [@jost_left_2021]
- Evidence for partisanship as a social identity
- We will use "identity-based" as a shorthand for double standards based on assigned categories such as class, race, and gender and "ideology-based" as a shorthand for double standards based on system-justification and political orientation.

<!-- I could move the last paragraph of the previous section to here. -->

@@

Growing evidence shows that, in many situations, observers apply a more lenient standards when judging moral transgressions by ingroup members [@abrams_double_2013; @endevelt_everyone_2021]. And yet, our findings suggest that, unlike judgments in other domains, judgments about collective action are not subject to double standards based on group identities such as class, race, or gender. 
<!-- Why did we not find any evidence for identity-based double standards? -->

@@

>>>>

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In five studies (total *N* = 2,979), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In three preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own ideological position---but showed no consistent ingroup bias in terms of class, race, and gender. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
keywords: collective action, double standards, social identity theory, system justification theory, Black Lives Matter
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: apa-template.tex
---

@@

Partisan differences can result from two social psychological processes. On the one hand, political partisanship reflects ideological differences in resistance to social change and acceptance of inequality which, in turn, are motivated by epistemic, existential, and relational needs [@jost_political_2003; @jost_left_2021]. As motivated cognition, ideological orientations shape how we think and feel about issues, actions, and events. On the other hand, political partisanship functions as a social identity [@finkel_political_2020; @osborne_psychology_2022] with evidence suggesting that partisans' policy preferences depends more on the stated position of their party than policy content [@cohen_party_2003]. Like other identities, political identities result in group-serving biases in how we evaluate ingroup and outgroup members. In practice, the two processes are often difficult to disentangle.

For now, we use "identity-based" to refer to double standards based on assigned social categories such as class, race, and gender and "ideology-based" to refer to double standards based on ideological alignment. We compare the evidence for identity and ideology processes in the General Discussion.

@@

Growing evidence shows that, in many situations, observers apply a more lenient standards when judging moral transgressions by ingroup members [@abrams_double_2013; @endevelt_everyone_2021]. And yet, our findings suggest that, unlike judgments in other domains, judgments about collective action are not subject to double standards based on group identities such as class, race, or gender.

@@

<!-- Instead, I should offer three explanations for the observed partisan biases: (1) ideology (Jost, 2020, 2021), (2) identity, and (3) moralizaton. I could close by suggesting experiments to tease them apart. Assymmetry is evidence against identity-based processes as we would expect ingroup bias in both directions. -->

### 2023-04-26 Documents/manuscript.tex
add paragraph on partisanship as a social identity

\noindent Collective action is a powerful force driving social change
but often sparks contention about what actions are acceptable means to
effect social change. In five studies (total \emph{N} = 2,979), we
investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In three
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police and for or against
restricting abortion in the United States. We found evidence for
ideology-based, but not for identity-based, double standards:
Participants judged the same protest actions to be more acceptable when
the protesters' cause aligned with their own ideological position---but
showed no consistent ingroup bias when judging collective action by
ingroup and outgroup members. Our findings have theoretical and
practical implications for understanding the often divided response to
prominent social movements.

@@

\begin{itemize}
\tightlist
\item
  Are Partisan Double Standards Based in Ideology or Identity?
\item
  Evidence for partisanship as a elective affinity (Jost, 2021)
\item
  Evidence for partisanship as a social identity
\item
  We will use ``identity-based'' as a shorthand for double standards
  based on assigned categories such as class, race, and gender and
  ``ideology-based'' as a shorthand for double standards based on
  system-justification and political orientation.
\end{itemize}

@@

@@

>>>>

\noindent Collective action is a powerful force driving social change
but often sparks contention about what actions are acceptable means to
effect social change. In five studies (total \emph{N} = 2,979), we
investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In three
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police and for or against
restricting abortion in the United States. We found evidence for
ideology-based, but not for identity-based, double standards:
Participants judged the same protest actions to be more acceptable when
the protesters' cause aligned with their own ideological position---but
showed no consistent ingroup bias in terms of class, race, and gender.
Our findings have theoretical and practical implications for
understanding the often divided response to prominent social movements.

@@

Partisan differences can result from two social psychological processes.
On the one hand, political partisanship reflects ideological differences
in resistance to social change and acceptance of inequality which, in
turn, are motivated by epistemic, existential, and relational needs
(Jost et al., 2003; Jost, 2021). As motivated cognition, ideological
orientations shape how we think and feel about issues, actions, and
events. On the other hand, political partisanship functions as a social
identity (Finkel et al., 2020; Harris et al., 2022) with evidence
suggesting that partisans' policy preferences depends more on the stated
position of their party than policy content (Cohen, 2003). Like other
identities, political identities result in group-serving biases in how
we evaluate ingroup and outgroup members. In practice, the two processes
are often difficult to disentangle.

For now, we use ``identity-based'' to refer to double standards based on
assigned social categories such as class, race, and gender and
``ideology-based'' to refer to double standards based on ideological
alignment. We compare the evidence for identity and ideology processes
in the General Discussion.

@@

\leavevmode\vadjust pre{\hypertarget{ref-cohen_party_2003}{}}%
Cohen, G. L. (2003). Party over policy: The dominating impact of group
influence on political beliefs. \emph{Journal of Personality and Social
Psychology}, \emph{85}(5), 808--822.
\url{https://doi.org/10.1037/0022-3514.85.5.808}

@@

\leavevmode\vadjust pre{\hypertarget{ref-osborne_psychology_2022}{}}%
Harris, E., Pärnamets, P., Sternisko, A., Robertson, C., \& Van Bavel,
J. J. (2022). The psychology and neuroscience of partisanship. In D.
Osborne \& C. G. Sibley (Eds.), \emph{The {Cambridge} handbook of
political psychology} (pp. 50--67). Cambridge University Press.
\url{https://doi.org/10.1017/9781108779104.005}
